---
title: "Taller sobre construcción de bases de datos"
author:
- name: Matías Frugone-Álvarez, Ph.D.
  url: https://matfrugone.netlify.app
  affiliation: Tecnico de datos Upwell
  affiliation_url: https://www.upwell.cl/web/
- name: Eugenia Gayo
  url: http://www.upwell.cl/web/biogeoquimica-de-humboldt/
  affiliation: Directora del proyecto Upwell
  affiliation_url: https://www.upwell.cl/web/investigadores/
- name: Carola Flores
  url: http://www.upwell.cl/web/interaccion-humano-ambiente/
  affiliation: Directora alterna del proyecto Upwell
  affiliation_url: https://www.upwell.cl/web/investigadores/
- name: Bernardo Broitman

  url: http://www.upwell.cl/web/ecologia-de-humboldt/
  affiliation: Investigador responsable Ecología de Humboldt
  affiliation_url: http://www.upwell.cl/web/ecologia-de-humboldt/
date: "`r Sys.Date()`"
description: |
  Bienvenidos a la website del taller de datos UPWELL. `Espero que lo disfruten!`
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```


## Bienvenidos  

Gracias por asistir a este curso de uso, manipulación y análisis de baes de datos del proyecto UPWELL 

```{r, include=FALSE, fig.cap="[Slides](slides/index.html)"}
knitr::include_url("slides/index.html")
```


\

## Pre-requisitos

### Esencial

+ Instalar rstudio.cloud <https://rstudio.cloud>.
+ Instalar la ultima version de [Zoom](https://zoom.us/download).
+ Join a pre-course technology check session - details are in the pre-course email.
+ ....

### Opcional

Please tune into class with a laptop that has the following installed:


<aside>
On the day of the class, we'll provide you with a link to an RStudio Cloud project that contains all of the course materials. 
</aside>

* Una version reciente de R (>=3.9.0), disponible en <https://cran.r-project.org/>

* Una version reciente de RStudio Desktop (>=1.3.0), disponible en <https://www.rstudio.com/download> (RStudio Desktop Open Source License)

* TLos R packages que vamos a usar;

    ```{r echo=TRUE, eval = FALSE}
    install.packages(c("tidyverse", "neotoma"))
    ```

<aside>
We will use the software listed above only as an important backup should there be problems with the classroom server connection.
</aside>

I look forward to meeting you,

Matías






\

## Links

+ [RStudio Cloud workspace](http://bit.ly/tmv-rmed2020)

### Slides
